DSCI 445 Final Report

### 2025-12-16 Paper.Rmd
paper polishing

## Motivation

Bank telemarketing campaigns are a major channel through which financial institutions promote term deposit products. Thsee marketing campaigns come with a cost and are often inefficient due to low client response rates. 
Prior work by Moro et al. (2014) discover that client characteristics and campaign related factors strongly influence subscription outcomes. Motivating a data-driven approach to improve targeting and decision-making in bank marketing. 
In their approach a Portuguese retail bank was studied using campaign data collected between 2008 and 2013.

Our dataset was a publicily available dataset, thus had less variables for privacy concerns of the bank customers.
The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns. These campaigns were done to encourage clients to subscribe to a long-term deposit product. Telemarketing is an expensive and time-consuming process; therefore, the bank would want to know in advance which customers are most likely to subscribe before any more investment takes place.
The dataset consists of sixteen variables describing demographic, financial and campaign related variables that enlist individual clients contacted during direct marketing efforts. A list of the 16 variable found below...

@@

### Problem Statement     
This study is motivated by the availability of bank marketing datasets that enable comparison of machine learning methods for predicting term deposit subscription behavior.

@@

In the following chpater of Introduction to Statistical Learning focuses on model selection and regularization techniques to improve predictive performance and reduce overfitting. It introduces methods such as subset selection, shrinkage (Ridge and LASSO regression), and dimension reduction to assess and enhance model accuracy while controlling variance. These approaches allow practitioners to identify parsimonious models that generalize well to new data, providing complementary strategies to traditional test error estimation. We apply these concepts to the models we fit on the bank data, to reduce overfitting and ultimately improve the generalization of our models to new clients. \

We start with a logistic regression (described in methodology section), then build up adjustments and regularizations to the logestic regression to make our baseline. We then move to another interpretable model: KNN. This model is considered transperent in a unique, due to it's neighboring factor.
Finally we fit a Boosting model. This state-of-the-art model is not considered interpretable. Although, there are methods used to extract the contributing variables in such models. We included this model is used to gage how much accucracy we are sacrificing for interpretability.

The authors employed time ordered data splitting rolling window evaluation. For feature enrichment, they implimented a semi-automatic approach including external intuitive knowledge from domain experts (bank manager) to optimize features. They compared Logistic regression with Decission trees, SVMs, and Neural networks; all complex 'black-box' models. While Moro et al. prioritized maximizing classification performance, our approach also examines model simplicity, automatic feature selection, and diagnostic checks for appropriate classifications (from confusion matrices).

## Methodology
For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets. Our models do not apply time-series modeling, since our data does not include sequintial variables. Although variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns, we do not model this data as a time series. This keeps the focus on cross-sectional relationships between client characteristics and subscription behavior.

@@

@@

```{r}
tran.coef
```

@@

Regularization techniques such as Ridge ($\ell_2$ penalty) and LASSO ($\ell_1$ penalty) are employed to reduce variance and perform feature selection, particularly in the presence of correlated predictors. LASSO (Least Absolute Shrinkage and Selection Operator) regression uses an $\ell_1$ penalty ($\alpha=1$). A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (Chapter 6). The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`. Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.

@@

We evaluated five main models from the course: ***#only 3 models***

@@

@@

### Model Training, Formulation, and Validation
All models were trained and evaluated within a consistent validation framework to ensure fair comparison and reliable estimation of out-of-sample performance. An 80/20 stratified train–test split was used, and all modeling decisions, including feature selection and hyperparameter tuning, were made using only the training data. This prevented information leakage and ensured that the test set provided an honest evaluation of generalization performance.

For the logistic regression models, feature selection was performed using backward stepwise selection based on AIC to obtain a parsimonious set of predictors. Residual diagnostics were used to assess model assumptions, leading to the inclusion of nonlinear terms where appropriate. Regularization techniques, including Ridge and LASSO regression, were applied to address multicollinearity and reduce variance. The regularization parameter was selected using 5-fold cross-validation.

@@

## Results

@@

## Discussion

@@

>>>>

# Motivation

Bank telemarketing campaigns are a major channel through which financial institutions promote term deposit products. These marketing campaigns come with a cost and are often inefficient due to low client response rates. 
Prior work by Moro et al. (2014) discover that client characteristics and campaign related factors strongly influence subscription outcomes. Motivating a data-driven approach to improve targeting and decision-making in bank marketing. 
In their approach a Portuguese retail bank was studied using campaign data collected between 2008 and 2013.

Our dataset was a publicly available dataset, thus had less variables for privacy concerns of the bank customers.
The Bank Marketing dataset was collected by a Portuguese banking institution during a series of direct telemarketing campaigns. These campaigns were done to encourage clients to subscribe to a long-term deposit product. Telemarketing is an expensive and time-consuming process; therefore, the bank would want to know in advance which customers are most likely to subscribe before any more investment takes place.
The dataset consists of sixteen variables describing demographic, financial and campaign related variables that enlist individual clients contacted during direct marketing efforts. A list of the 16 variable found below...

@@

## Problem Statement     
This study is motivated by the availability of bank marketing datasets that enable comparison of machine learning methods for predicting term deposit subscription behavior.

@@

In the following chapter of Introduction to Statistical Learning focuses on model selection and regularization techniques to improve predictive performance and reduce overfitting. It introduces methods such as subset selection, shrinkage (Ridge and LASSO regression), and dimension reduction to assess and enhance model accuracy while controlling variance. These approaches allow practitioners to identify parsimonious models that generalize well to new data, providing complementary strategies to traditional test error estimation. We apply these concepts to the models we fit on the bank data, to reduce overfitting and ultimately improve the generalization of our models to new clients. \

We start with a logistic regression (described in methodology section), then build up adjustments and regularization to the logestic regression to make our baseline. We then move to another interpretable model: KNN. This model is considered transparent in a unique, due to it's neighboring factor.
Finally we fit a Boosting model. This state-of-the-art model is not considered interpretable. Although, there are methods used to extract the contributing variables in such models. We included this model is used to gage how much accuracy we are sacrificing for interpretability.

The authors in (Moro et al.) employed time ordered data splitting rolling window evaluation. For feature enrichment, they implemented a semi-automatic approach including external intuitive knowledge from domain experts (bank manager) to optimize features. They compared Logistic regression with Decision trees, SVMs, and Neural networks; all complex 'black-box' models. While (Moro et al.) prioritized maximizing classification performance, our approach also examines model simplicity, automatic feature selection, and diagnostic checks for appropriate classifications (from confusion matrices).

# Methodology
For each of our models we implement a k-fold CV approach, with k=5 training/test split to maintain the distribution of the response variable across the sets. Our models do not apply time-series modeling, since our data does not include sequential variables. Although variables such as campaign month or day are available within the dataset, which can be considered to hold temporal patterns, we do not model this data as a time series. This keeps the focus on cross-sectional relationships between client characteristics and subscription behavior. An 80/20 random train–test split was used, and all modeling decisions, including feature selection and hyper-parameter tuning, were made using only the training data. This prevented information leakage and ensured that the test set provided an honest evaluation of generalization performance.

@@

### Interpretability
To understand the meaning of the coefficients, we can transform the formula into the log-odds (logit) form:
$$ \log\left(\frac{P(Y=1 \mid \mathbf{x})}{1 - P(Y=1 \mid \mathbf{x})}\right) = \beta_0 + \beta_1 x_1 + \dots + \beta_k x_k$$
Consider a single coefficient, say $\beta_{balance}$, which relates to a client's average bank account balance. If the estimated coefficient for balance is $\beta_{balance} = 0.001$, this means that a one-unit increase in the client's balance is associated with a change of 0.001 in the log-odds of subscribing to a term deposit, assuming all other variables remain constant. This change in log-odds translates to a multiplicative change of $e^{0.001} \approx 1.001$ in the odds. Therefore, for every unit increase in balance, the odds of subscription increase by a factor of 1.001. This shows a clear relationship between the coefficient and the target variable.

Without loss of generality, we can generalize this interpretation and estimation process to logistic regression models with multiple variables and their variations, such as incorporating Lasso regularization (for variable selection and complexity control) or applying polynomial transformations to the predictor variables (to capture non-linear relationships).

### Model fit

@@

```{r, echo=FALSE}
full_coefs <- summary(glm.tran)$coefficients
ballance_coefs <- c("poly(balance, 4)1", "poly(balance, 4)2", "poly(balance, 4)3", "poly(balance, 4)4")
subset_coefs <- full_coefs[ballance_coefs, c(1, 4)]

print(subset_coefs)
```

@@

Regularization techniques such as Ridge ($\ell_2$ penalty) and LASSO ($\ell_1$ penalty) are employed to reduce variance and perform feature selection, particularly in the presence of correlated predictors. LASSO (Least Absolute Shrinkage and Selection Operator) regression uses an $\ell_1$ penalty ($\alpha=1$). A key property of the LASSO is that it forces the coefficients of some variables to be exactly zero, thereby performing automatic feature selection (ISLR Chapter 6). The LASSO plot similarly uses cross-validation to select the optimal $\lambda_{\min}$, `r round(lambda.min.lasso, 5)`. Unlike Ridge, the LASSO is expected to yield a more parsimonious model by excluding irrelevant predictors entirely.

@@

We evaluated five main models from the course:

2. **K-Nearest Neighbors (KNN)**

@@

3. **Boosting**

@@

### Model Training, Formulation, and Validation

@@

# Results

@@

# Discussion

@@

James, Gareth, et al. An Introduction to Statistical Learning : With Applications in R. Springer, 2013.

Moro, Sérgio, et al. A Data-Driven Approach to Predict the Success of Bank Telemarketing. Decision Support Systems, vol. 62, June 2014, pp. 22–31.
